---
title: "NM Microbiome - 02 Susceptibility Analysis"
author: "Orr Tobaly & Nikol Elyashov"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

## 1. Setup and Library Loading

```{r load-libraries}
# Load required packages
library(phyloseq)
library(ANCOMBC)
library(dplyr)
library(ggplot2)
library(tidyr)
library(readxl)
library(DT)
library(patchwork)
library(microbiome)
library(viridis)

# Source custom functions
source("../R/preprocessing_functions.R")
source("../R/differential_abundance.r")
source("../R/visualization_functions.r")

# Set data paths
processed_data_dir <- "../data/processed/"
results_dir <- "../results/"

# Create results directory if it doesn't exist
if(!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}
```

## 2. Load Preprocessed Data

```{r load-data}
# Load the preprocessed phyloseq object
ps <- readRDS(file.path(processed_data_dir, "preprocessed_phyloseq.rds"))

# Load the microbiome dataset for additional functionality if needed
dataset <- readRDS(file.path(processed_data_dir, "preprocessed_microbiome_dataset.rds"))

# Check that our carriage classifications are present
print("Carriage group distribution:")
table(sample_data(ps)$carriage_group)

print("Susceptibility group distribution:")
table(sample_data(ps)$susceptibility_group)

# Import demographic data for additional covariates
demographic_data <- read_excel("../data/raw/demographic data _Jan2022_final.YM.20241230.xlsx")

# Clean and prepare demographic data
demographic_data <- demographic_data %>%
  mutate(
    pid = as.character(pid),  # Ensure pid is character for joining
    boardingschool = factor(boardingschool),
    selfsmoking = factor(selfsmoking),
    secondsmoker = factor(secondsmoker),
    AB_month = factor(AB_month),  # Antibiotic use in previous month
    AB_year = factor(AB_year)     # Antibiotic use in previous year
  )

# Merge relevant demographic variables with sample data
# First, get the current sample data
sample_data_df <- data.frame(sample_data(ps))

# Ensure 'pid' is preserved as the identifier
sample_data_with_demo <- sample_data_df %>%
  left_join(demographic_data %>% 
              select(pid, boardingschool, selfsmoking, secondsmoker, 
                     AB_month, AB_year, yearofbirth, family_members), 
            by = "pid")

# Rebuild the sample_data ensuring original sample names are preserved
rownames(sample_data_with_demo) <- sample_data_df$SampleID

# Update the phyloseq object with new sample data
sample_data(ps) <- sample_data(sample_data_with_demo)

# Verify the update worked
print("Updated sample data columns:")
names(sample_data(ps))
```

## 3. Analysis Design and Data Subset Selection

This analysis focuses on identifying microbiome signatures associated with susceptibility to NM acquisition. We'll examine differences between individuals who acquired NM at some point ("Ever_acquired_NM") versus those who consistently remained non-carriers ("Never_acquired_NM").

```{r subset-selection}
# We'll focus on baseline samples (time point 0) to identify initial microbiome differences
ps_baseline <- subset_samples(ps, time == "0")

# Create susceptibility analysis subset
ps_suscept <- subset_samples(ps_baseline, 
                            susceptibility_group %in% c("Ever_acquired_NM", "Never_acquired_NM"))

# Check sample sizes for our analysis
print("Sample counts for susceptibility analysis:")
print(table(sample_data(ps_suscept)$susceptibility_group))

# Verify that we have adequate data for comparison
if(min(table(sample_data(ps_suscept)$susceptibility_group)) < 10) {
  warning("Warning: One of the susceptibility groups has fewer than 10 samples!")
}
```

## 4. Exploratory Analysis of Susceptibility Groups

Before conducting differential abundance analysis, we'll explore alpha and beta diversity patterns and overall taxonomic composition across our susceptibility groups.

```{r exploratory-analysis}
# Plot alpha diversity by susceptibility group
alpha_div_metrics <- c("Shannon", "Simpson", "Observed")
alpha_plots <- list()

for(metric in alpha_div_metrics) {
  alpha_plots[[metric]] <- plot_richness(ps_suscept, x = "susceptibility_group", measures = metric) +
    geom_boxplot(alpha = 0.6) +
    labs(title = paste(metric, "Diversity by Susceptibility Group"),
         x = "Susceptibility Group") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Combine alpha diversity plots
wrap_plots(alpha_plots, ncol = 2)

# Beta diversity analysis (PCoA) by susceptibility group
ord <- ordinate(ps_suscept, method = "PCoA", distance = "bray")
beta_plot <- plot_ordination(ps_suscept, ord, color = "susceptibility_group") +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse() +
  labs(title = "PCoA of microbiome composition by susceptibility group",
       color = "Susceptibility Group") +
  theme_minimal()
print(beta_plot)

# PERMANOVA test for statistical significance of beta diversity differences
permanova_result <- vegan::adonis2(
  phyloseq::distance(ps_suscept, method = "bray") ~ 
    phyloseq::sample_data(ps_suscept)$susceptibility_group, 
  permutations = 999
)
print("PERMANOVA results:")
print(permanova_result)

# Taxonomic composition by susceptibility group
ps_genus <- tax_glom(ps_suscept, taxrank = "Genus")
top_genera <- names(sort(taxa_sums(ps_genus), decreasing = TRUE))[1:15]
ps_top_genera <- prune_taxa(top_genera, ps_genus)

# Convert to relative abundance
ps_top_genera_rel <- transform_sample_counts(ps_top_genera, function(x) x / sum(x))

# Plot taxonomic composition
taxa_comp_plot <- plot_bar(ps_top_genera_rel, fill = "Genus", x = "susceptibility_group") +
  labs(title = "Relative abundance of top 15 genera by susceptibility group",
       x = "Susceptibility Group", 
       y = "Relative Abundance") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
print(taxa_comp_plot)
```

## 5. Differential Abundance Analysis with ANCOM-BC2

We'll use ANCOM-BC2 to identify differentially abundant taxa between susceptibility groups, while accounting for potential confounding factors.

```{r ancom-bc2-analysis}
# Set random seed for reproducibility
set.seed(123)

# Run ANCOM-BC2 using our custom function
ancombc2_result <- run_ancombc2(
  ps_object = ps_suscept,
  tax_level = "Genus",
  formula_str = "susceptibility_group + selfsmoking + boardingschool + AB_month",
  group_var = "susceptibility_group",
  prv_cut = 0.025,
  seed = 123
)

# Save the results
saveRDS(ancombc2_result, file = file.path(results_dir, "ancombc2_susceptibility_results.rds"))

# Extract basic results to check
res_primary <- ancombc2_result$res
print(paste("Total number of taxa analyzed:", nrow(res_primary)))

# Get the column names related to log fold changes and differential abundance
lfc_cols <- grep("^lfc_", colnames(res_primary), value = TRUE)
diff_cols <- grep("^diff_", colnames(res_primary), value = TRUE)
print("Comparisons in the results:")
print(gsub("^lfc_", "", lfc_cols))

# Count the number of significant taxa for each comparison
for(i in seq_along(diff_cols)) {
  comparison <- gsub("^diff_", "", diff_cols[i])
  n_sig <- sum(res_primary[[diff_cols[i]]] == 1)
  n_sig_with_ss <- sum(res_primary[[diff_cols[i]]] == 1 & 
                       res_primary[[paste0("passed_ss_", comparison)]] == TRUE)
  
  print(paste0("Comparison ", comparison, ":"))
  print(paste0("  - Significant taxa: ", n_sig))
  print(paste0("  - Significant taxa passing sensitivity score: ", n_sig_with_ss))
}
```

## 6. Visualization of Significant Taxa

```{r visualize-results}
# Visualize ANCOM-BC2 results using our custom function
ancombc2_vis <- visualize_ancombc2_results(
  ancombc2_result,
  title = "Susceptibility Analysis"
)

# Display the plots
if(length(ancombc2_vis$plots) > 0) {
  for(plot_name in names(ancombc2_vis$plots)) {
    print(ancombc2_vis$plots[[plot_name]])
  }
} else {
  print("No significant taxa found to visualize.")
}

# Display a summary table of significant taxa
display_results <- function(vis_results, title) {
  cat("\n====== ", title, " ======\n")
  
  # Check if there are significant results
  if(length(vis_results$sig_taxa) == 0) {
    cat("No significant differential abundance detected.\n")
    return(NULL)
  }
  
  # Display results for each comparison
  for(comparison in names(vis_results$sig_taxa)) {
    sig_taxa <- vis_results$sig_taxa[[comparison]]
    
    if(nrow(sig_taxa) > 0) {
      cat("\nComparison:", comparison, "\n")
      cat("Number of significant taxa:", nrow(sig_taxa), "\n")
      
      # Create a data frame for display
      result_df <- sig_taxa %>% 
        select(taxon, lfc) %>%
        arrange(desc(abs(lfc)))
      
      # Display the data frame
      print(result_df)
    }
  }
}

# Display the results summary
display_results(ancombc2_vis, "SIGNIFICANT TAXA FOR SUSCEPTIBILITY ANALYSIS")

# Save significant taxa results to CSV
for(comparison in names(ancombc2_vis$sig_taxa)) {
  if(length(ancombc2_vis$sig_taxa[[comparison]]) > 0 && nrow(ancombc2_vis$sig_taxa[[comparison]]) > 0) {
    write.csv(
      ancombc2_vis$sig_taxa[[comparison]],
      file = file.path(results_dir, paste0("sig_taxa_", comparison, ".csv")),
      row.names = FALSE
    )
  }
}

# Create a heatmap of significant taxa abundances
create_abundance_heatmap <- function(ps_obj, sig_taxa_names, group_var) {
  # Check if we have significant taxa
  if(length(sig_taxa_names) == 0) {
    return(NULL)
  }
  
  # Subset to significant taxa
  ps_subset <- prune_taxa(sig_taxa_names, ps_obj)
  
  # Transform to relative abundance
  ps_rel <- transform_sample_counts(ps_subset, function(x) x / sum(x))
  
  # Extract abundance data
  abund_mat <- as(otu_table(ps_rel), "matrix")
  if(taxa_are_rows(ps_rel)) {
    abund_mat <- t(abund_mat)
  }
  
  # Get sample metadata
  sample_meta <- data.frame(sample_data(ps_rel))
  group_data <- sample_meta[[group_var]]
  
  # Create a data frame for ggplot
  abund_df <- as.data.frame(abund_mat)
  abund_df$SampleID <- rownames(abund_df)
  abund_df$Group <- group_data
  
  # Reshape to long format
  abund_long <- abund_df %>%
    pivot_longer(cols = -c(SampleID, Group),
                names_to = "Taxon",
                values_to = "Abundance")
  
  # Create heatmap
  heatmap_plot <- ggplot(abund_long, aes(x = SampleID, y = Taxon, fill = Abundance)) +
    geom_tile() +
    facet_grid(. ~ Group, scales = "free_x", space = "free_x") +
    scale_fill_viridis_c(name = "Relative\nAbundance") +
    labs(title = paste("Abundance of Significant Taxa by", group_var),
         x = "Sample", y = "Taxon") +
    theme_minimal() +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
  
  return(heatmap_plot)
}

# Extract significant taxa across all comparisons
all_sig_taxa <- unique(unlist(lapply(ancombc2_vis$sig_taxa, function(x) x$taxon)))

# Create and display heatmap if we have significant taxa
if(length(all_sig_taxa) > 0) {
  heatmap_plot <- create_abundance_heatmap(ps_suscept, all_sig_taxa, "susceptibility_group")
  if(!is.null(heatmap_plot)) {
    print(heatmap_plot)
    ggsave(file.path(results_dir, "susceptibility_sig_taxa_heatmap.png"), 
           heatmap_plot, width = 12, height = 8)
  }
}
```

## 7. Covariate Importance Analysis

In addition to susceptibility status, we included several covariates in our model. Here we'll examine their associations with taxa abundance.

```{r covariate-importance}
# Extract covariate results from ANCOM-BC2
covariate_results <- list()
covariate_cols <- c("selfsmoking", "boardingschool", "AB_month")

for(covar in covariate_cols) {
  # Get relevant columns for this covariate
  lfc_col <- grep(paste0("^lfc_", covar), colnames(res_primary), value = TRUE)
  diff_col <- grep(paste0("^diff_", covar), colnames(res_primary), value = TRUE)
  
  if(length(lfc_col) > 0 && length(diff_col) > 0) {
    # Filter for significant taxa associated with this covariate
    sig_taxa <- res_primary %>%
      filter(!!sym(diff_col) == 1) %>%
      select(taxon, !!sym(lfc_col)) %>%
      rename(lfc = !!sym(lfc_col)) %>%
      arrange(desc(abs(lfc)))
    
    covariate_results[[covar]] <- sig_taxa
    
    # Print summary
    cat("\nSignificant taxa associated with", covar, ":", nrow(sig_taxa), "\n")
    if(nrow(sig_taxa) > 0) {
      print(head(sig_taxa, 10))
    }
  }
}

# Visualize significant taxa for each covariate
for(covar in names(covariate_results)) {
  if(nrow(covariate_results[[covar]]) > 0) {
    # Limit to top 15 taxa by absolute log fold change
    top_n_taxa <- min(nrow(covariate_results[[covar]]), 15)
    plot_data <- covariate_results[[covar]][1:top_n_taxa, ]
    
    # Create bar plot
    p <- ggplot(plot_data, aes(x = reorder(taxon, lfc), y = lfc, 
                             fill = lfc > 0)) +
      geom_bar(stat = "identity") +
      scale_fill_manual(values = c("#619CFF", "#F8766D"), 
                      name = "Direction", 
                      labels = c("Decreased", "Increased")) +
      labs(title = paste("Taxa associated with", covar),
           x = "Taxon",
           y = "Log Fold Change") +
      coord_flip() +
      theme_minimal()
    
    print(p)
    
    # Save the plot
    ggsave(file.path(results_dir, paste0("covariate_", covar, "_taxa.png")), 
           p, width = 10, height = 6)
  }
}
```

## 8. Interpretation and Biological Significance

In this section, we'll interpret the biological significance of the differentially abundant taxa identified in our analysis.

```{r biological-interpretation}
# Extract the names of significant taxa that passed sensitivity score filter
sig_taxa_passing_ss <- list()

for(comparison in names(ancombc2_vis$sig_taxa)) {
  sig_taxa_df <- ancombc2_vis$sig_taxa[[comparison]]
  if(!is.null(sig_taxa_df) && nrow(sig_taxa_df) > 0) {
    # Get taxon names, assuming they're already filtered for significance and sensitivity
    sig_taxa_passing_ss[[comparison]] <- sig_taxa_df$taxon
  }
}

# Get the full taxonomy information for significant taxa
get_taxa_table <- function(ps_obj, taxa_names) {
  if(length(taxa_names) == 0) return(NULL)
  
  # Extract tax table
  tax_table <- as.data.frame(tax_table(ps_obj))
  
  # Filter for significant taxa
  sig_tax_table <- tax_table[rownames(tax_table) %in% taxa_names, ]
  
  # Add taxon name as a column
  sig_tax_table$taxon <- rownames(sig_tax_table)
  
  return(sig_tax_table)
}

# Print taxonomic information for significant taxa
for(comparison in names(sig_taxa_passing_ss)) {
  cat("\n=== Taxonomy of significant taxa for", comparison, "===\n")
  taxa_names <- sig_taxa_passing_ss[[comparison]]
  
  if(length(taxa_names) > 0) {
    taxa_table <- get_taxa_table(ps_suscept, taxa_names)
    if(!is.null(taxa_table)) {
      print(taxa_table)
    }
  } else {
    cat("No significant taxa passing sensitivity score filter\n")
  }
}

# Summarize key findings
cat("\n=== Key findings from susceptibility analysis ===\n")

# Count total significant taxa
total_sig_taxa <- unique(unlist(sig_taxa_passing_ss))
cat("Total unique significant taxa identified:", length(total_sig_taxa), "\n")

# Highlight taxonomic patterns
if(length(total_sig_taxa) > 0) {
  all_tax_table <- get_taxa_table(ps_suscept, total_sig_taxa)
  if(!is.null(all_tax_table)) {
    # Summarize at Phylum level
    phylum_summary <- all_tax_table %>%
      group_by(Phylum) %>%
      summarise(count = n()) %>%
      arrange(desc(count))
    
    cat("\nPhylum distribution of significant taxa:\n")
    print(phylum_summary)
    
    # Summarize at Family level
    family_summary <- all_tax_table %>%
      group_by(Family) %>%
      summarise(count = n()) %>%
      arrange(desc(count))
    
    cat("\nFamily distribution of significant taxa:\n")
    print(head(family_summary, 10))
  }
} else {
  cat("No significant taxa to summarize taxonomic patterns\n")
}
```

## 9. Export Results and Visualizations

```{r export-results}
# Save the full ANCOM-BC2 results object
saveRDS(ancombc2_result, file = file.path(results_dir, "susceptibility_ancombc2_results.rds"))

# Save the visualization results
saveRDS(ancombc2_vis, file = file.path(results_dir, "susceptibility_visualization_results.rds"))

# Create a summary of significant findings
susceptibility_summary <- list(
  analysis_info = list(
    date = Sys.Date(),
    samples_analyzed = nrow(sample_data(ps_suscept)),
    group_sizes = table(sample_data(ps_suscept)$susceptibility_group),
    covariates = c("selfsmoking", "boardingschool", "AB_month")
  ),
  significance_summary = list(
    total_taxa_analyzed = nrow(res_primary),
    significant_taxa = lapply(diff_cols, function(col) sum(res_primary[[col]] == 1)),
    significant_taxa_passing_ss = lapply(diff_cols, function(col) {
      ss_col <- gsub("diff_", "passed_ss_", col)
      sum(res_primary[[col]] == 1 & res_primary[[ss_col]] == TRUE)
    })
  ),
  differentially_abundant_taxa = sig_taxa_passing_ss
)

# Save the summary
saveRDS(susceptibility_summary, 
       file = file.path(results_dir, "susceptibility_analysis_summary.rds"))

# Save a text summary of results
sink(file = file.path(results_dir, "susceptibility_results_summary.txt"))
cat("=== NM SUSCEPTIBILITY ANALYSIS RESULTS ===\n\n")
cat("Analysis Date:", as.character(Sys.Date()), "\n")
cat("Samples Analyzed:", nrow(sample_data(ps_suscept)), "\n")
cat("Group Sizes:\n")
print(table(sample_data(ps_suscept)$susceptibility_group))
cat("\n")

cat("=== SIGNIFICANT TAXA ===\n\n")
display_results(ancombc2_vis, "SIGNIFICANT TAXA FOR SUSCEPTIBILITY ANALYSIS")
cat("\n")

cat("=== PERMANOVA RESULTS ===\n\n")
print(permanova_result)
cat("\n")

sink()
```

## 10. Session Info

```{r session-info}
sessionInfo()
```
