---
title: "NM Microbiome - 02 Original Susceptibility Analysis"
author: "Orr Tobaly & Nikol Elyashov"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

## 1. Setup and Library Loading

```{r load-libraries}
# Load required packages
library(phyloseq)
library(ANCOMBC)
library(dplyr)
library(ggplot2)
library(tidyr)
library(readxl)
library(DT)
library(patchwork)
library(microbiome)
library(viridis)

# Source custom functions
source("../R/preprocessing_functions.R")
source("../R/differential_abundance.r")
source("../R/visualization_functions.r")

# Set data paths
processed_data_dir <- "../data/processed/"
results_dir <- "../results/original_susceptibility/"

# Create results directory if it doesn't exist
if(!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}
```

## 2. Load Preprocessed Data

```{r load-data}
# Load the preprocessed phyloseq object
ps <- readRDS(file.path(processed_data_dir, "preprocessed_phyloseq.rds"))

# Load the microbiome dataset for additional functionality if needed
dataset <- readRDS(file.path(processed_data_dir, "preprocessed_microbiome_dataset.rds"))

# Check that our carriage classifications are present
print("Carriage group distribution:")
table(sample_data(ps)$carriage_group)

# Import demographic data for additional covariates
demographic_data <- read_excel("../data/raw/demographic data _Jan2022_final.YM.20241230.xlsx")

# Clean and prepare demographic data
demographic_data <- demographic_data %>%
  mutate(
    pid = as.character(pid),  # Ensure pid is character for joining
    boardingschool = factor(boardingschool),
    selfsmoking = factor(selfsmoking),
    secondsmoker = factor(secondsmoker),
    AB_month = factor(AB_month),  # Antibiotic use in previous month
    AB_year = factor(AB_year)     # Antibiotic use in previous year
  )

# Merge relevant demographic variables with sample data
# First, get the current sample data
sample_data_df <- data.frame(sample_data(ps))

# Ensure 'pid' is preserved as the identifier
sample_data_with_demo <- sample_data_df %>%
  left_join(demographic_data %>% 
              select(pid, boardingschool, selfsmoking, secondsmoker, 
                     AB_month, AB_year, yearofbirth, family_members), 
            by = "pid")

# Rebuild the sample_data ensuring original sample names are preserved
rownames(sample_data_with_demo) <- sample_data_df$SampleID

# Update the phyloseq object with new sample data
sample_data(ps) <- sample_data(sample_data_with_demo)

# Verify the update worked
print("Updated sample data columns:")
names(sample_data(ps))
```

## 3. Analysis Design and Data Subset Selection

Unlike the expanded susceptibility analysis, here we focus on the original carriage group classification. We'll examine differences between individuals who started as non-carriers and acquired NM at a later time point ("Acquisition_carrier") versus those who remained non-carriers throughout the study ("Non_carrier").

```{r subset-selection}
# We'll focus on baseline samples (time point 0) to identify initial microbiome differences
ps_baseline <- subset_samples(ps, time == "0")

# Create original susceptibility analysis subset - focusing on acquisition vs non-carriers
ps_suscept_original <- subset_samples(ps_baseline, 
                                     carriage_group %in% c("Acquisition_carrier", "Non_carrier"))

# Check sample sizes for our analysis
print("Sample counts for original susceptibility analysis:")
print(table(sample_data(ps_suscept_original)$carriage_group))

# Verify that we have adequate data for comparison
if(min(table(sample_data(ps_suscept_original)$carriage_group)) < 10) {
  warning("Warning: One of the carriage groups has fewer than 10 samples. Results may be unstable.")
} else {
  print("Both groups have adequate sample sizes (≥10). Analysis should be robust.")
}
```

## 4. Exploratory Analysis of Original Carriage Groups

Before conducting differential abundance analysis, we'll explore alpha and beta diversity patterns and overall taxonomic composition across our carriage groups.

```{r exploratory-analysis}
# Plot alpha diversity by carriage group
alpha_div_metrics <- c("Shannon", "Simpson", "Observed")
alpha_plots <- list()

for(metric in alpha_div_metrics) {
  alpha_plots[[metric]] <- plot_richness(ps_suscept_original, x = "carriage_group", measures = metric) +
    geom_boxplot(alpha = 0.6) +
    labs(title = paste(metric, "Diversity by Carriage Group"),
         x = "Carriage Group") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Combine alpha diversity plots
wrap_plots(alpha_plots, ncol = 2)

# Beta diversity analysis (PCoA) by carriage group
ord <- ordinate(ps_suscept_original, method = "PCoA", distance = "bray")
beta_plot <- plot_ordination(ps_suscept_original, ord, color = "carriage_group") +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse() +
  labs(title = "PCoA of microbiome composition by carriage group",
       color = "Carriage Group") +
  theme_minimal()
print(beta_plot)

# PERMANOVA test for statistical significance of beta diversity differences
permanova_result <- vegan::adonis2(
  phyloseq::distance(ps_suscept_original, method = "bray") ~ 
    phyloseq::sample_data(ps_suscept_original)$carriage_group, 
  permutations = 999
)
print("PERMANOVA results:")
print(permanova_result)

# Taxonomic composition by carriage group
ps_genus <- tax_glom(ps_suscept_original, taxrank = "Genus")
top_genera <- names(sort(taxa_sums(ps_genus), decreasing = TRUE))[1:15]
ps_top_genera <- prune_taxa(top_genera, ps_genus)

# Convert to relative abundance
ps_top_genera_rel <- transform_sample_counts(ps_top_genera, function(x) x / sum(x))

# Plot taxonomic composition
taxa_comp_plot <- plot_bar(ps_top_genera_rel, fill = "Genus", x = "carriage_group") +
  labs(title = "Relative abundance of top 15 genera by carriage group",
       x = "Carriage Group", 
       y = "Relative Abundance") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
print(taxa_comp_plot)
```

## 5. Differential Abundance Analysis with ANCOM-BC2

We'll use ANCOM-BC2 to identify differentially abundant taxa between the original carriage groups, while accounting for potential confounding factors.

```{r ancom-bc2-analysis}
# Set random seed for reproducibility
set.seed(123)


# Run function directly
ancombc2_result <- ancombc2(
  data = ps_suscept_original,
  tax_level = "Genus",
  fix_formula = "carriage_group + cohort + selfsmoking + boardingschool + AB_month",
  rand_formula = NULL,
  p_adj_method = "holm",
  pseudo_sens = TRUE,
  prv_cut = 0.025,
  lib_cut = 1000,
  s0_perc = 0.1,
  group = "carriage_group",
  struc_zero = TRUE,
  neg_lb = TRUE,
  alpha = 0.05,
  n_cl = 2,
  global = FALSE,
  pairwise = TRUE,
  verbose = TRUE
)

# # Run ANCOM-BC2 using our custom function, now using carriage_group
# ancombc2_result <- run_ancombc2(
#   ps_object = ps_suscept_original,
#   tax_level = "Genus",
#   formula_str = "carriage_group + selfsmoking + boardingschool + AB_month",
#   group_var = "carriage_group",
#   prv_cut = 0.025,
#   seed = 123
# )

# Save the results
saveRDS(ancombc2_result, file = file.path(results_dir, "ancombc2_original_susceptibility_results.rds"))

# Extract basic results to check
res_primary <- ancombc2_result$res
print(paste("Total number of taxa analyzed:", nrow(res_primary)))

# Get the column names related to log fold changes and differential abundance
lfc_cols <- grep("^lfc_", colnames(res_primary), value = TRUE)
diff_cols <- grep("^diff_", colnames(res_primary), value = TRUE)
print("Comparisons in the results:")
print(gsub("^lfc_", "", lfc_cols))

# Count the number of significant taxa for each comparison
for(i in seq_along(diff_cols)) {
  comparison <- gsub("^diff_", "", diff_cols[i])
  n_sig <- sum(res_primary[[diff_cols[i]]] == 1)
  
  ss_col <- paste0("passed_ss_", comparison)
  if(ss_col %in% colnames(res_primary)) {
    n_sig_with_ss <- sum(res_primary[[diff_cols[i]]] == 1 & 
                         res_primary[[ss_col]] == TRUE)
    
    print(paste0("Comparison ", comparison, ":"))
    print(paste0("  - Significant taxa: ", n_sig))
    print(paste0("  - Significant taxa passing sensitivity score: ", n_sig_with_ss))
  } else {
    print(paste0("Comparison ", comparison, ": ", n_sig, " significant taxa"))
  }
}
```

## 6. Visualization of Significant Taxa

```{r visualize-results}
# Visualize ANCOM-BC2 results using our custom function
ancombc2_vis <- visualize_ancombc2_results(
  ancombc2_result,
  title = "Original Susceptibility Analysis"
)

# Display the plots
if(length(ancombc2_vis$plots) > 0) {
  for(plot_name in names(ancombc2_vis$plots)) {
    print(ancombc2_vis$plots[[plot_name]])
    
    # Save the plots
    ggsave(file.path(results_dir, paste0("diff_abundant_taxa_", plot_name, ".png")), 
           ancombc2_vis$plots[[plot_name]], width = 10, height = 8)
  }
} else {
  print("No significant taxa found to visualize.")
}

# Display a summary table of significant taxa
display_results <- function(vis_results, title) {
  cat("\n====== ", title, " ======\n")
  
  # Check if there are significant results
  if(length(vis_results$sig_taxa) == 0) {
    cat("No significant differential abundance detected.\n")
    return(NULL)
  }
  
  # Display results for each comparison
  for(comparison in names(vis_results$sig_taxa)) {
    sig_taxa <- vis_results$sig_taxa[[comparison]]
    
    if(nrow(sig_taxa) > 0) {
      cat("\nComparison:", comparison, "\n")
      cat("Number of significant taxa:", nrow(sig_taxa), "\n")
      
      # Create a data frame for display
      result_df <- sig_taxa %>% 
        select(taxon, lfc) %>%
        arrange(desc(abs(lfc)))
      
      # Display the data frame
      print(result_df)
    }
  }
}

# Display the results summary
display_results(ancombc2_vis, "SIGNIFICANT TAXA FOR ORIGINAL SUSCEPTIBILITY ANALYSIS")

# Save significant taxa results to CSV
for(comparison in names(ancombc2_vis$sig_taxa)) {
  if(length(ancombc2_vis$sig_taxa[[comparison]]) > 0 && nrow(ancombc2_vis$sig_taxa[[comparison]]) > 0) {
    write.csv(
      ancombc2_vis$sig_taxa[[comparison]],
      file = file.path(results_dir, paste0("sig_taxa_", comparison, ".csv")),
      row.names = FALSE
    )
  }
}

# Extract significant taxa across all comparisons
all_sig_taxa <- unique(unlist(lapply(ancombc2_vis$sig_taxa, function(x) x$taxon)))

# Create a heatmap of significant taxa abundances
if(length(all_sig_taxa) > 0) {
  # First, create a tax-glom'ed version of the phyloseq object at the genus level
  ps_genus <- tax_glom(ps_suscept_original, taxrank = "Genus")
  
  # Function to map between ANCOM-BC2 taxa names and phyloseq taxa
  create_taxa_mapping <- function(ps_obj, ancombc2_taxa) {
    # Extract tax table
    tax_df <- as.data.frame(tax_table(ps_obj))
    
    # Create a mapping dictionary
    taxa_map <- list()
    
    # Process each ANCOM-BC2 taxon name
    for (taxon_name in ancombc2_taxa) {
      # Parse the taxon name to extract level and value
      if (grepl(":", taxon_name)) {
        # Format is "Level:Name"
        parts <- strsplit(taxon_name, ":")[[1]]
        level <- parts[1]
        value <- parts[2]
        
        # Find matching taxa in the phyloseq object
        if (level %in% colnames(tax_df)) {
          matching_indices <- which(tax_df[[level]] == value)
          if (length(matching_indices) > 0) {
            taxa_map[[taxon_name]] <- rownames(tax_df)[matching_indices]
          }
        }
      } else {
        # Try to find a direct match
        if (taxon_name %in% rownames(tax_df)) {
          taxa_map[[taxon_name]] <- taxon_name
        }
      }
    }
    
    return(taxa_map)
  }
  
  # Function to create a heatmap using the mapped taxa
  create_mapped_heatmap <- function(ps_obj, ancombc2_taxa, group_var) {
    # Get taxa mapping
    taxa_mapping <- create_taxa_mapping(ps_obj, ancombc2_taxa)
    
    # Check if we found any matches
    if (length(taxa_mapping) == 0) {
      warning("No matching taxa found in the phyloseq object")
      return(NULL)
    }
    
    message(sprintf("Found %d matching taxa out of %d significant taxa", 
                    length(taxa_mapping), length(ancombc2_taxa)))
    
    # Flatten the mapping to get all matching taxa IDs
    all_matching_taxa <- unique(unlist(taxa_mapping))
    
    # Subset the phyloseq object to only include matching taxa
    ps_subset <- prune_taxa(all_matching_taxa, ps_obj)
    
    # Transform to relative abundance
    ps_rel <- transform_sample_counts(ps_subset, function(x) x / sum(x))
    
    # Extract abundance data
    abund_mat <- as(otu_table(ps_rel), "matrix")
    if (taxa_are_rows(ps_rel)) {
      abund_mat <- t(abund_mat)
    }
    
    # Get the taxonomy information for labeling
    tax_info <- as.data.frame(tax_table(ps_rel))
    
    # Create row labels using genus and family
    row_labels <- apply(tax_info, 1, function(x) {
      if (!is.na(x["Genus"]) && x["Genus"] != "") {
        return(x["Genus"])
      } else if (!is.na(x["Family"]) && x["Family"] != "") {
        return(paste0(x["Family"], " (unknown genus)"))
      } else {
        return(rownames(tax_info)[which(rownames(tax_info) == rownames(tax_info)[which(tax_info[,1] == x[1])])[1]])
      }
    })
    names(row_labels) <- rownames(tax_info)
    
    # Get sample metadata
    sample_meta <- data.frame(sample_data(ps_rel))
    group_data <- sample_meta[[group_var]]
    
    # Create a data frame for ggplot
    abund_df <- as.data.frame(abund_mat)
    abund_df$SampleID <- rownames(abund_df)
    abund_df$Group <- group_data
    
    # Reshape to long format
    abund_long <- abund_df %>%
      pivot_longer(cols = -c(SampleID, Group),
                  names_to = "Taxon",
                  values_to = "Abundance") %>%
      mutate(TaxonLabel = row_labels[Taxon])
    
    # Create heatmap
    heatmap_plot <- ggplot(abund_long, aes(x = SampleID, y = TaxonLabel, fill = Abundance)) +
      geom_tile() +
      facet_grid(. ~ Group, scales = "free_x", space = "free_x") +
      scale_fill_viridis_c(name = "Relative\nAbundance") +
      labs(title = paste("Abundance of Significant Taxa by", group_var),
           x = "Sample", y = "Taxon") +
      theme_minimal() +
      theme(axis.text.x = element_blank(),
            axis.ticks.x = element_blank())
    
    return(heatmap_plot)
  }
  
  # Create and display the heatmap
  heatmap_plot <- create_mapped_heatmap(ps_genus, all_sig_taxa, "carriage_group")
  if (!is.null(heatmap_plot)) {
    print(heatmap_plot)
    ggsave(file.path(results_dir, "significant_taxa_heatmap.png"), 
           heatmap_plot, width = 12, height = 8)
  } else {
    message("Could not create heatmap - no matching taxa found")
  }
} else {
  message("No significant taxa found to display in heatmap")
}
```

## 7. Covariate Importance Analysis

In addition to carriage status, we included several covariates in our model. Here we'll examine their associations with taxa abundance.

```{r covariate-importance}
# Extract covariate results from ANCOM-BC2
covariate_results <- list()
covariate_cols <- c("selfsmoking", "boardingschool", "AB_month")

for(covar in covariate_cols) {
  # Get relevant columns for this covariate
  lfc_col <- grep(paste0("^lfc_", covar), colnames(res_primary), value = TRUE)
  diff_col <- grep(paste0("^diff_", covar), colnames(res_primary), value = TRUE)
  
  if(length(lfc_col) > 0 && length(diff_col) > 0) {
    # Filter for significant taxa associated with this covariate
    sig_taxa <- res_primary %>%
      filter(!!sym(diff_col) == 1) %>%
      select(taxon, !!sym(lfc_col)) %>%
      rename(lfc = !!sym(lfc_col)) %>%
      arrange(desc(abs(lfc)))
    
    covariate_results[[covar]] <- sig_taxa
    
    # Print summary
    cat("\nSignificant taxa associated with", covar, ":", nrow(sig_taxa), "\n")
    if(nrow(sig_taxa) > 0) {
      print(head(sig_taxa, 10))
    }
  }
}

# Visualize significant taxa for each covariate
for(covar in names(covariate_results)) {
  if(nrow(covariate_results[[covar]]) > 0) {
    # Limit to top 15 taxa by absolute log fold change
    top_n_taxa <- min(nrow(covariate_results[[covar]]), 15)
    plot_data <- covariate_results[[covar]][1:top_n_taxa, ]
    
    # Create bar plot
    p <- ggplot(plot_data, aes(x = reorder(taxon, lfc), y = lfc, 
                             fill = lfc > 0)) +
      geom_bar(stat = "identity") +
      scale_fill_manual(values = c("#619CFF", "#F8766D"), 
                      name = "Direction", 
                      labels = c("Decreased", "Increased")) +
      labs(title = paste("Taxa associated with", covar),
           x = "Taxon",
           y = "Log Fold Change") +
      coord_flip() +
      theme_minimal()
    
    print(p)
    
    # Save the plot
    ggsave(file.path(results_dir, paste0("covariate_", covar, "_taxa.png")), 
           p, width = 10, height = 6)
  }
}
```

## 8. Interpretation and Biological Significance

In this section, we'll interpret the biological significance of the differentially abundant taxa identified in our analysis.

```{r biological-interpretation}
# Extract the names of significant taxa that passed sensitivity score filter
sig_taxa_passing_ss <- list()

for(comparison in names(ancombc2_vis$sig_taxa)) {
  sig_taxa_df <- ancombc2_vis$sig_taxa[[comparison]]
  if(!is.null(sig_taxa_df) && nrow(sig_taxa_df) > 0) {
    # Get taxon names, assuming they're already filtered for significance and sensitivity
    sig_taxa_passing_ss[[comparison]] <- sig_taxa_df$taxon
  }
}

# Get the full taxonomy information for significant taxa
get_taxa_table <- function(ps_obj, taxa_names) {
  if(length(taxa_names) == 0) return(NULL)
  
  # Extract tax table
  tax_table <- as.data.frame(tax_table(ps_obj))
  
  # Filter for significant taxa
  sig_tax_table <- tax_table[rownames(tax_table) %in% taxa_names, ]
  
  # Add taxon name as a column
  sig_tax_table$taxon <- rownames(sig_tax_table)
  
  return(sig_tax_table)
}

# Print taxonomic information for significant taxa
for(comparison in names(sig_taxa_passing_ss)) {
  cat("\n=== Taxonomy of significant taxa for", comparison, "===\n")
  taxa_names <- sig_taxa_passing_ss[[comparison]]
  
  if(length(taxa_names) > 0) {
    taxa_table <- get_taxa_table(ps_suscept_original, taxa_names)
    if(!is.null(taxa_table)) {
      print(taxa_table)
    }
  } else {
    cat("No significant taxa passing sensitivity score filter\n")
  }
}

# Summarize key findings
cat("\n=== Key findings from original susceptibility analysis ===\n")

# Count total significant taxa
total_sig_taxa <- unique(unlist(sig_taxa_passing_ss))
cat("Total unique significant taxa identified:", length(total_sig_taxa), "\n")

# Highlight taxonomic patterns
if(length(total_sig_taxa) > 0) {
  all_tax_table <- get_taxa_table(ps_suscept_original, total_sig_taxa)
  if(!is.null(all_tax_table)) {
    # Summarize at Phylum level
    phylum_summary <- all_tax_table %>%
      group_by(Phylum) %>%
      summarise(count = n()) %>%
      arrange(desc(count))
    
    cat("\nPhylum distribution of significant taxa:\n")
    print(phylum_summary)
    
    # Summarize at Family level
    family_summary <- all_tax_table %>%
      group_by(Family) %>%
      summarise(count = n()) %>%
      arrange(desc(count))
    
    cat("\nFamily distribution of significant taxa:\n")
    print(head(family_summary, 10))
  }
} else {
  cat("No significant taxa to summarize taxonomic patterns\n")
}
```

## 9. Comparison with Expanded Susceptibility Analysis

Here we'll compare results from our original carriage group analysis with the expanded susceptibility analysis to see if there are consistent taxa signatures.

```{r compare-analyses}
# Check if the expanded analysis results exist
expanded_results_file <- file.path("../results", "ancombc2_susceptibility_results.rds")
if(file.exists(expanded_results_file)) {
  # Load the expanded analysis results
  expanded_results <- readRDS(expanded_results_file)
  expanded_res_primary <- expanded_results$res
  
  # Extract significant taxa from expanded analysis
  expanded_sig_taxa <- list()
  exp_diff_cols <- grep("^diff_", colnames(expanded_res_primary), value = TRUE)
  
  for(diff_col in exp_diff_cols) {
    comparison <- gsub("^diff_", "", diff_col)
    ss_col <- paste0("passed_ss_", comparison)
    
    if(ss_col %in% colnames(expanded_res_primary)) {
      sig_taxa <- expanded_res_primary$taxon[
        expanded_res_primary[[diff_col]] == 1 & 
        expanded_res_primary[[ss_col]] == TRUE
      ]
    } else {
      sig_taxa <- expanded_res_primary$taxon[expanded_res_primary[[diff_col]] == 1]
    }
    
    expanded_sig_taxa[[comparison]] <- sig_taxa
  }
  
  # Compare with current results
  cat("\n=== Comparison with Expanded Susceptibility Analysis ===\n")
  
  # Find overlap in significant taxa
  all_expanded_sig_taxa <- unique(unlist(expanded_sig_taxa))
  
  if(length(total_sig_taxa) > 0 && length(all_expanded_sig_taxa) > 0) {
    overlap_taxa <- intersect(total_sig_taxa, all_expanded_sig_taxa)
    
    cat("Total significant taxa in original analysis:", length(total_sig_taxa), "\n")
    cat("Total significant taxa in expanded analysis:", length(all_expanded_sig_taxa), "\n")
    cat("Number of overlapping taxa:", length(overlap_taxa), "\n")
    
    if(length(overlap_taxa) > 0) {
      cat("\nOverlapping taxa:\n")
      overlap_tax_table <- get_taxa_table(ps_suscept_original, overlap_taxa)
      if(!is.null(overlap_tax_table)) {
        print(overlap_tax_table %>% select(Phylum, Class, Order, Family, Genus, taxon))
      }
    }
  } else {
    cat("No overlapping significant taxa found between the analyses\n")
  }
} else {
  cat("Expanded analysis results not found. Skipping comparison.\n")
}
```

## 10. Export Results and Visualizations

```{r export-results}
# Save the full ANCOM-BC2 results object
saveRDS(ancombc2_result, file = file.path(results_dir, "original_susceptibility_ancombc2_results.rds"))

# Save the visualization results
saveRDS(ancombc2_vis, file = file.path(results_dir, "original_susceptibility_visualization_results.rds"))

# Create a summary of significant findings
susceptibility_summary <- list(
  analysis_info = list(
    date = Sys.Date(),
    samples_analyzed = nrow(sample_data(ps_suscept_original)),
    group_sizes = table(sample_data(ps_suscept_original)$carriage_group),
    covariates = c("selfsmoking", "boardingschool", "AB_month")
  ),
  significance_summary = list(
    total_taxa_analyzed = nrow(res_primary),
    significant_taxa = lapply(diff_cols, function(col) sum(res_primary[[col]] == 1)),
    significant_taxa_passing_ss = lapply(diff_cols, function(col) {
      ss_col <- gsub("diff_", "passed_ss_", col)
      if(ss_col %in% colnames(res_primary)) {
        sum(res_primary[[col]] == 1 & res_primary[[ss_col]] == TRUE)
      } else {
        sum(res_primary[[col]] == 1)
      }
    })
  ),
  differentially_abundant_taxa = sig_taxa_passing_ss
)

# Save the summary
saveRDS(susceptibility_summary, 
       file = file.path(results_dir, "original_susceptibility_analysis_summary.rds"))

# Save a text summary of results
sink(file = file.path(results_dir, "original_susceptibility_results_summary.txt"))
cat("=== NM ORIGINAL SUSCEPTIBILITY ANALYSIS RESULTS ===\n\n")
cat("Analysis Date:", as.character(Sys.Date()), "\n")
cat("Samples Analyzed:", nrow(sample_data(ps_suscept_original)), "\n")
cat("Group Sizes:\n")
print(table(sample_data(ps_suscept_original)$carriage_group))
cat("\n")

cat("=== SIGNIFICANT TAXA ===\n\n")
display_results(ancombc2_vis, "SIGNIFICANT TAXA FOR ORIGINAL SUSCEPTIBILITY ANALYSIS")
cat("\n")

cat("=== PERMANOVA RESULTS ===\n\n")
print(permanova_result)
cat("\n")

sink()
```

## 11. Session Info

```{r session-info}
sessionInfo()
```